Title

### 2024-06-24 paper/main.tex
updates

\section{Background}
\label{sec:background}
We represent uni-dimensional variables as $ X $ and multi-dimensional variables
as $ \bm{Z} $. We consider the case of estimating association between $ X $,
and $ Y $ is the presence of a set of conditioning variables $ \bm{Z} = Z_1,
\cdots, Z_k $, potentially being an empty set $ \bm{Z} = \emptyset $. We use $
\rvert \bm{Z} \rvert $ to denote the cardinality of the variable $ \bm{Z} $. We
consider various scenarios where the variables $ X $, $ Y $ and $ \bm{Z} $
could be either discrete, ordinal, or continuous. With mixed data, we mean an
arbitrary mix of variable types among $ X $, $ Y $, and $ \bm{Z} $.

@@

\subsection{Measures of Association}
As our DAG construction procedure relies on a measure of conditional
association between variables, in this section we list out some of the already
known measures for specific data types. These measure of associations are
typically the effect size of various conditional indepdnence tests but they
don't need to be. However, there are no generalized measure of association for
mixed data, that we propose in the next section. This is by no means an
exhaustive list, rather just to give an idea of what can be done.

@@

\section{A Measure of Association for Mixed Data}
\label{sec:mixed_association}

In this section, we develop a (conditional) measure of association for mixed
data. This measure is similar to partial correlation in the sense that it uses
a residualization approach and takes the association between the residuals.

Given two variables $ X $ and $ Y $, and a set of conditioanl variables $
\bm{Z} $, we are interested in computing the conditional association, $ \rho(X,
Y; Z) $. Similar to partial correlation, we start by computing the residuals
for $ X $ and $ Y $ using the $ \bm{Z} $ as the predictor varaibles and then
compute the assocation between the residuals. As the residuals for mixed
variable is not straightforward to define, we use the mixed data
residualization approach presented in Ankan and Textor 2023.

Given a variable $ X $ with predictor variables $ \bm{Z} $, we first train an
estimator, $ E_X $ for predicting $ X $ using $ \bm{Z} $. If the varialbes are
ordinal or categorical, the estimator needs to be probability estimators for
example a Random Forest Estimator or XGBoost. Given an estimator we can compute
the residuals as follows:
\begin{enumerate}
	\item If $ X $ is continuous, we simply take the difference between the true and estimated value.
		$$ R_X = X - E_X(\bm{Z}) $$
	\item If $ X $ is ordinal, we take the Li-Sheperd residual.
		$$ R_X = E_X(\bm{Z}) $$
	\item If $ X $ is categorical, we one hot encode the variable and Li-Sheperd residuals as well.
		$$ R_X = E_X(\bm{Z}) $$
\end{enumerate}
As a result of this residualization approach, the residuals are a matrix of
continuous values. The shape of this matrix depends on the type of variable. If
the variables are continuous or ordinal, we end up with a residual matrix of
size $ n \times 1 $. If the the variable is categorical we end up with a
residual matrix of size $ n \times (k-1) $, where $ k $ is the cardinality of
the categorical variables. We have dropped one of the one hot encoded columns
because it is redundant.

Next we need a measure to quantify the association between these residuals. One
of the general ways to have assocation between sets of variables is to use
canonical correlations. Canonical correlations are generalization of
correlations to multi-dimensional variables.

@@

Multiple measures of association based on canonical correlation are available such as:
\begin{itemize}
	\item Wilks' Lambda: $ \Lambda = \prod (1 - \phi_i^2) $
	\item Roy's Largest Root: $ \theta = \phi_{\max}^2 $
	\item Pillai's Trace: $ \tau = \sum \phi_i^2 $
\end{itemize}

@@

\begin{enumerate}
	\item Is bounded between $ 0 $ and $ 1 $.
	\item Is independent of the sample size.
	\item Is indifferent to the one-hot encoding scheme used for categorical variables.
	\item In the case of both continuous variables, reduces to partial correlation.
\end{enumerate}

\subsection{Connections to other measures}
In the continuous variable case, this is equivalent to partial correlation.

This natually is a generalization of partial correlations. Furthermore, it is 
also a generalization of discrete case.

>>>>

\section{Background}
\label{sec:background}
We represent uni-dimensional variables as $ X $ and multi-dimensional variables
as $ \bm{Z} $. All the variables are considered to be any mix of continuous,
ordinal, or categorical unless specified explicitly.

We consider the case of estimating association between $ X $ and $ Y $ is the
presence of a set of conditioning variables $ \bm{Z} = Z_1, \cdots, Z_k $,
potentially being an empty set $ \bm{Z} = \emptyset $. We use $ \rvert \bm{Z}
\rvert $ to denote the cardinality of the set of variables $ \bm{Z} $. We
consider various scenarios where the variables $ X $, $ Y $ and $ \bm{Z} $
could be either discrete, ordinal, or continuous. With mixed data, we mean an
arbitrary mix of variable types among $ X $, $ Y $, and $ \bm{Z} $.

@@

\subsection{Measures of (Conditional) Association}
As our DAG construction procedure relies on a measure of conditional
association between variables, in this section we list out some of the already
known measures for specific data types. These measure of associations are
typically the effect size of various conditional indepdnence tests but they
don't need to be. However, there are no generalized measure of association for
mixed data, that we propose in the next section. This is by no means an
exhaustive list, rather just to give an idea of what can be done.

@@

\section{Measure of Association for Mixed Data}
\label{sec:mixed_association}

In this section, we develop a novel (conditional) measure of association for
mixed data. The presented measure extends partial correlation to mixed data by
combining a mixed-data residualization method with a multivariate measure of
association.

Given two variables $ X $ and $ Y $, and a set of conditional variables $
\bm{Z} $, we are interested in computing the conditional association, $ \rho(X,
Y; Z) $. As our measure is based on residulaization, we start by computing the
residuals for $ X $ and $ Y $ using the $ \bm{Z} $ as the predictor variables.
We use the mixed data residualization approach from \citet{Ankan2023}. We
consider each variable type separately and repeat the following steps for both
$ X $ and $ Y $:

\begin{enumerate}
	\item $ X $ is continuous: We start by training a regression model, $
		E_X: X \sim \bm{Z} $. Next we take the predictions from this
		model $ \hat{x} = E_X(\bm{Z}) $ and lastly the residuals are
		computed as the difference between the true and the predicted
		values.
		$$ R_X = X - \hat{x} $$
	\item $ X $ is ordinal: We start by training a probability estimator, $
		p_X: X \sim \bm{Z} $, and take probability predictions from it
		$ \hat{p}_X: p_X(\bm{Z}) $. Then we compute the residuals as follows:
		$$ R_X = $$
	\item $ X $ is categorical: We again start by training a probability estimator
		$ p_X: X \sim \bm{Z} $, and take the probability estimates from it $ \hat{p}_X: p_Z(\bm{Z}) $.
		Next, we one-hot encode the categorical variable and then compute the residuals
		as follows:
		$$ R_X $$
\end{enumerate}

Here, users can choose the type of estimator that they want to use. There are
single estimators such as Random Forest or XGBoost that can work for both the
probability estimates and regression estimates. Or users can also choose
different estimators such as variants of linear regression for computing the
residuals. After the above steps are repeated for both $ X $ and $ Y $, we end
up with two residual matrices: $ R_X $ and $ R_Y $. The type of the variable
determines the shape for these matrices, i.e., if the variable is continuous we
get a residual matrix of shape $ ( n \times 1 ) $ and if the variable is
categorical we have a residual matrix of shape $ ( n \times (k-1)) $ (where one
of the dummy encoded variables is dropped to avoid multicollinearity)

Now, we have the two residuals and we need a measure to quantify the
association between these residuals. As the residuals are matrices, we can
treat them as sets of random variables, and in multivariate analysis canonical
correlations have been used to quantify association between two sets of
variables. Canonical correlations are generalization of correlations to
multi-dimensional variables where they try to find linear combinations of the
given sets of variables that maximize the correlation between them.

@@

Many measures of association based on canonical correlation are also available such as:
\begin{itemize}
	\item Wilks' Lambda: $ \Lambda = \prod (1 - \phi_i^2) $
	\item Roy's Largest Root: $ \theta = \phi_{\max}^2 $
	\item Pillai's Trace: $ \tau = \sum \phi_i^2 $
\end{itemize}

@@

\begin{enumerate}
	\item Is bounded between $ 0 $ and $ 1 $, with $ 0 $ representing no association and $ 1 $ representing a perfect linear relationship.
	\item Is independent of the sample size.
	\item As canonical correlation finds linear transformations, the measure is independent of the one-hot encoding scheme used for categorical variables.
	\item In the case of both continuous variables, it is equivalent to partial correlations.
	\item For both discrete variables, the measure is connected to Cramers\' V, a well know effect size measure for discrete variables.
\end{enumerate}
